# Go Deeper {#GoDeeper}

This chapter tells how to get more out of your uBiome results.

From [uBiome Explorer](https://explorer.ubiome.com) click on the "Insights" tab ( Figure \@ref(fig:explorerInsightsPage)) which may be more informative for beginners.If this is your first time here, go ahead and breeze through this page, but just ignore most of the conclusions. As we'll learn, scientists have barely scratched the surface of what we will eventually understand about the microbiome, so think of this page as a teaser -- a hint of the potential. If your results show something you didn't expect, say on diversity or body weight, don't worry: it almost certainly just means that the Insights pane is wrong about you. Later we'll see how to pull out far more interesting and useful results.

```{r explorerInsightsPage, fig.cap = "The Insights tab provides a high-level view of what uBiome thinks is most actionable about your microbiome."}
knitr::include_graphics(file.path(getwd(),"assets/ubiomeExplorerInsightsPageMar29.jpg"))
```

The one part of the Insights page that you _should_ take seriously is the section called "most uncommon bacteria". Let's jump to that section now.

## How are you different?

Since so little about the microbiome is well-understood, most of the time it will be hard to interpret your results by themselves. This is where the power of uBiome's citizen science can help, because the web page lets you easily compare your results with other people.  

```{r explorerCompare, fig.cap = "The Compare view shows the differences between my sample and others in the uBiome database."}
knitr::include_graphics(file.path(getwd(),"assets/ubiomeExplorerCompareMar29.jpg"))
```


In this case, I can see that my sample has about 64% of the abundance of _Firmicutes_ as other people in the uBiome database. You can sort this field by abundance (ratio) and click on the various taxonomic categories to dig deeper into your microbiome and find microbes that are particularly rare in your sample. (Figure \@ref(fig:explorerRare))

```{r explorerRare, fig.cap = "My sample shows very low levels of these microbes, compared to the hundreds of thousands of others in the uBiome database."}
knitr::include_graphics(file.path(getwd(),"assets/ubiomeExploreRareMar29.jpg"))
```

For example, I found that I have much less _Prevotella_ and _Butyvibrio_, two common microbes than the vast majority of uBiome customers.  I also found much higher-than-normal levels of _Deinococcus-Thermus_, an extremophile usually only found in harsh environments, like the impossibly acidic human stomach. Why would I have so many more than other people like me?

One clue is the similarly-high levels of Proteobacteria, which as we just noted is often high in people with gastro-intestinal issues. Maybe I was exposed to something that released a bit of stomach acid, and its extremophile _Deinococcus_ into my colon.

_Verrucomicrobia_ is elevated too, though not by much. This is a Phylum that is generally associated with soil microbes, though it's not unusual to find it in the gut as well.

I looked back at my notes during the period of a few days before this sample, and interestingly I had been doing work in my garden at the time. We have deer and other animals that come near our house...is it possible that I was exposed to some animal excrement, maybe some of which somehow made it into my gut?

It's just a theory, and perhaps an unlikely one, but I wonder if my body was fighting some type of pathogen as a result of my outdoor activities. Now before we get too excited, keep in mind that I never felt the slightest bit ill during this period. Can I thank my microbiome? Perhaps I was exposed to something nasty, the kind of bug that under other circumstances might have wreaked havoc and left me terribly sick for days. Now I have no way of knowing for sure, but this tidbit makes me wonder if maybe my microbiome acted as a guardian angel, protecting me from something much worse.

Similarly, your own results comparing to others may lead you to clues about aspects of your environment you didn't otherwise suspect.

Incidentally, since this test I've sent dozens of additional samples to uBiome and never saw that _Deinococcus_ thing again. Good riddance? or maybe I should be thankful for the protection, and hope that little bits of it may still be present, undetectably in my stomach, waiting to fight another external enemy the next time I come in contact with something in the soil.
