## The Oral Microbiome* {#startMouth}

```{r}
# This is now read in loadMHGData.r
# mouth.genus <- subset_samples(sprague.genus.norm, Site == "mouth")
# mouth.species <- subset_samples(sprague.species.norm, Site == "mouth")
```


The first place that microbes enter is also one of the richest and most variable environments in the body.


Scientists added an odorless compound from wine to a culture of known oral bacteria, and sure enough: the bacteria generated compounds that we can smell: terpenes, benzenic compounds and lipid derivatives. Each of us has a unique oral microbiome, and scientists were able to that that this inter-person variability is large enough to explain at least some of the differences in how each of us perceive a glass of wine.^[@munoz-gonzalez_ability_2015]


What are the most important species in my mouth?

See the Appendix for an [overall summary of my mouth experiments](#appendixMyMouth)




How much variability is there within the mouth? To find out, I took four samples immediately after waking up one morning. I was careful to swab only in a specific part of the mouth.

```{r mouthsetup1}

mouth.Nov6 <- subset_samples(mouth.genus, Date == "2016-11-06")
mouth.Nov6 <- prune_taxa(taxa_sums(mouth.Nov6)>42,mouth.Nov6)
sample_data(mouth.Nov6)$Label <- c("Lips","Tongue","Cheek (Right)", "Cheek (Left)")

```



```{r mouthshow}
kable(mhg_abundance(mouth.Nov6, colnames = "Label", top = 20)/10000, digits=2)
```

While there is some variation in the cheeks, there is a dramatic difference between them and the lips or tongue. Also interesting is the way the lips are dominated by just three taxa that make up more than 85% of the total abundance. In Inverse Simpson terms, the lips are the least diverse, whereas the tongue is the most diverse.


```{r}
plot_richness(mouth.Nov6, measure = "Shannon", x = "Label") + ggtitle("Mouth Microbiome Diversity")
```

Let's start with the genus level. How much *Streptococcus* has been in my mouth, and to the degree that we know at the species level, which types of species are there?

Here I find something alarming: a new type of _Streptococcus_ appears to have colonized my mouth, somewhat suddenly, around April of this year. _Streptococcus sp. BS35a_ seems to have been unknown to my mouth until then. Why? And is it really alarming?

I've noticed no change in my health, dental or otherwise, and although I'm doing experiments all the time I can't point to anything in particular that might have caused such a shift in that particular microbe.



```{r mouthStrep}
mouth.species <- subset_samples(sprague.species.norm, Site == "mouth" & Reads >10000 ) #& Date > "2016-07-01")
#as.character(tax_table(mouth.species)[grep("Strep*",tax_table(mouth.species))])

sample_data(sprague.all_ranks.norm)$Date <- as.Date(sample_data(sprague.all_ranks.norm)$Date, origin = "1899-12-30")
# plot_bar(subset_taxa(subset_samples(sprague.all_ranks.norm,
#                                     Site == "mouth" & Date > "2016-10-01"),
#                      grepl("Strep*",Species) ),
#          x = "Date", fill = "Species") + 
#   ggtitle("Sprague Mouth Microbes") + 
#   scale_x_date(date_labels = "%b20%y", date_breaks = "1 month", date_minor_breaks = "1 week") +
#   ylab("Abundance (%)") + scale_y_continuous(labels=function(x)x/10000) + ylab("Abundance (%)") +
#   theme(legend.position = "bottom")
```

At the species level, I eliminated all samples with under 10,000 reads.  We see something interesting: for no apparent reason, the species of _Streptococcus_ detected in my mouth has changed. Suddenly, in December 2016 my mouth was colonized by a particular species that had not been there before. Why? I hadn't done anything special; I'm not aware of any major lifestyle or other changes to cause this.


```{r}
# Make the X-axis ordered by date, but not as a date axis. 
# sample_data(mouth.species)$Dateorder <- as.character(sample_data(mouth.species)$Date) 
# plot_bar(subset_taxa(prune_taxa(taxa_sums(mouth.species)>50000, mouth.species),
#                      grepl("Strep*",Species) ),
#          x = "Dateorder", fill = "Species") + 
#   ggtitle("Sprague Mouth Microbes") + 
# #  scale_x_date(date_labels = "%b20%y", date_breaks = "1 month", date_minor_breaks = "1 week") +
#   ylab("Abundance (%)") + scale_y_continuous(labels=function(x)x/10000) + ylab("Abundance (%)") +
#   theme(legend.position = "right" ) #, axis.title.x = element_blank(),axis.text.x=element_blank())

sample_data(mouth.species)$Dateorder <- as.character(sample_data(mouth.species)$Date) 
plot_bar(subset_taxa(prune_taxa(taxa_sums(mouth.species)>50000, mouth.species),
                     grepl("Strep*",Species) ),
         x = "Dateorder", fill = "Species") + 
  ggtitle(paste("Sprague Mouth Microbes","through",tail(sample_data(mouth.species)$Date,1))) + theme(axis.text.x=element_blank()) + xlab("Dates (ordered)") + 
#  scale_x_date(date_labels = "%b20%y", date_breaks = "1 month", date_minor_breaks = "1 week") +
  ylab("Abundance (%)") + scale_y_continuous(labels=function(x)x/10000) + ylab("Abundance (%)") +
  theme(legend.position = "right" ) #, axis.title.x = element_blank(),axis.text.x=element_blank())


```

I confirmed with the lab that it's not contamination. 

What's _especially_ odd is that I experienced a shift like this *twice* now in one year. After comfortably floating along with Species _Streptococcus sp. BS35a_ for more than six months, suddenly in August the balance shifted again, this time to _Streptococcus sp. 11aTha1_. Will it shift again? Who knows?


