## Gut Issues

Probably the most common reason people are motivated to study their microbiome is due to their discomfort with their gut.

CB is a health-conscious woman who has long suffered from gut issues.

Let's start with a heatmap of all her gut samples to see if we can spot any patterns

```{r caseStudyCBExplore}

cb.gut.genus <- subset_samples(people.gut.norm, Username == "cb")
cb.gut.genus <- prune_taxa(taxa_sums(cb.gut.genus)>42, cb.gut.genus)
plot_heatmap(cb.gut.genus, sample.order = "Date", sample.label = "Date")
# ggplot(data = sample_data(cb.gut.genus)[,c("Date","Reads")], aes(x=Date,y=Reads)) + geom_bar(stat="identity") 


```

Hard to tell from this single case. How about if we look at another person ("NJ1")


First let's make a similarity plot it to see if we can spot something unusual:

```{r}
#plot_top_heat(nj1)
people.tmp <- people.family
sample_data(people.tmp)$Username <-ifelse(sample_data(people.tmp)$Username=="NJ1","NJ1","People")
make_ord_for(subset_samples(people.tmp, Site == "gut"))
```

Nothing special there. She seems firmly in the middle of most of the rest of the people.

Now let's look for specific microbes at the phylum level:

```{r}

nj1 <- subset_samples(people.norm, Username == "NJ1")
nj1 <- prune_taxa(taxa_sums(nj1)>42, nj1)
nj1.phylum <- subset_samples(people.phylum, Username == "NJ1")
nj1.phylum <- prune_taxa(taxa_sums(nj1.phylum)>42, nj1.phylum)

knitr::kable(mhg_abundance(nj1.phylum)/10000, digits = 2)

```

Nothing obviously different there than with others I've seen. Might be nice to see more Verrucomicrobia in the second (7-25) sample, but nothing pops as particularly special. How about the genus level:

```{r, results = "asis"}
knitr::kable(mhg_abundance(nj1)/10000, digits = 2)

```

The most striking finding is how different the second sample is from the first. Why?

When I look closer, it appears that much of the difference is driven by one genus, _Colinsella_, of which she has among the highest abundances I've seen. In my own gut samples, I've never had more than 5% of that genus, but she has more than 8%. Is that good or bad? Who knows? But it's something I would spend more time studying.


```{r}

mhg_density_for_taxa(people.norm, "Collinsella") +
  annotate(
  geom = "segment",
  color = "blue",
  x = as.numeric(otu_table(
  subset_samples(people.norm, Username == "NJ1")
  )["Collinsella"]) / 10000,
  xend = as.numeric(otu_table(
  subset_samples(people.norm, Username == "NJ1")
  )["Collinsella"]) / 10000 ,
  y = -Inf,
  yend = Inf
  )

           
```




Clearly different at nearly all taxa.  Which microbe disappeared in the second sample?

```{r}

mhg_unique(subset_samples(nj1, Date > "2017-07-01"), subset_samples(nj1, Date < "2017-07-01"))
```

and vice versa:

```{r}
kable(mhg_unique(subset_samples(nj1, Date < "2017-07-01"),subset_samples(nj1, Date > "2017-07-01"))[1]/10000, digits = 2, col.names = c("% Unique to first sample"))
```


Let's look at another case:  "Ed" is a German man on a mostly vegetarian diet.

```{r}

ed <- subset_samples(people.norm, Username == "AD" & Site == "gut")
ed.norm <- prune_taxa(taxa_sums(ed)>42, ed)

#plot_top_heat(ed.norm)

c.people <- people.norm # compare set
sample_data(c.people)$Username <- "Other"


ed.all <- subset_samples(people.phylum, Username == "AD")
ed.ssr <- sample_data(ed.all)$SSR
sample_data(c.people)$Username[sample_data(c.people)$SSR %in% ed.ssr] <- "You"


  
  
ed.phylum <- subset_samples(people.phylum, Username == "AD" & Site == "gut")
ed.phylum.norm <- prune_taxa(taxa_sums(ed.phylum)>42, ed.phylum)
mhg_abundance(ed.phylum.norm)/10000

mhg_abundance(ed.norm)/10000

# using another database: not for public
# load(paste0(DATA_DIR,"/ubiome corp/healthy/healthy897Genus.RData"))
# sample_data(healthy.ps)$Site <- "gut"
# healthyPlusEd <- merge_phyloseq(healthy.ps,ed.norm)
# ord <- ordinate(healthyPlusEd, method = "NMDS", distance = "bray" )
# plot_ordination(healthyPlusEd, ord,  color = "Username", shape = "Site")# label = "Label")


#ord <- ordinate(subset_samples(c.people, Site == "gut"), method = "DCA", distance = "bray" )

make_ord_for(subset_samples(c.people, Site == "gut"), color = "Username")

```

### Celiac Patients


L and her friend S suffer from the gluten disorder called Celiac disease.

```{r}
lori.phylum <- subset_samples(people.phylum,  Username == c("em1"))
lori  <- subset_samples(people.norm,  Username == c("em1"))
sandy.phylum <- subset_samples(people.phylum,  Username == c("em2"))
sandy <-subset_samples(people.norm,  Username == c("em2"))
mhg_abundance(merge_phyloseq(lori.phylum,sandy.phylum),top=6)/10000

mhg_abundance(merge_phyloseq(lori,sandy),top=10)/10000

#mhg_plot_top_heat(merge_phyloseq(lori,sandy))

```


I don't often see L's level of Fusobacterium. A quick check, though, shows that she's not the highest.

```{r}
mhg_density_for_taxa(people.norm,"Fusobacterium") + 
  annotate(
  geom = "segment",
  color = "blue",
  x = as.numeric(otu_table(
  subset_samples(people.norm, Username == "em1"))["Fusobacterium"] / 10000),
  xend = as.numeric(otu_table(
  subset_samples(people.norm, Username == "em1"))["Fusobacterium"] / 10000),
  y = -Inf,
  yend = Inf
  )

plot_richness(merge_phyloseq(subset_samples(people.family,Username=="em1"),subset_samples(people.family,Username=="em2")), measures=c("InvSimpson"))

ls.ssr <- c(sample_data(lori)$SSR,sample_data(sandy)$SSR)
  
  
c.people <- subset_samples(people.family, Site == "gut") # compare set
sample_data(c.people)$Username <- "Other"

#sample_data(c.people)$Username[sample_data(c.people)$SSR %in% ls.ssr] <- subset_samples(c.people)

sample_data(c.people)$Username[sample_data(c.people)$SSR %in% ls.ssr] <- sample_data(people.norm)$Username[sample_data(people.norm)$SSR %in% ls.ssr]

# loriSandyHealthy <- merge_phyloseq(lori,sandy,healthy.ps)
# ord <- ordinate(loriSandyHealthy, method = "NMDS", distance = "bray" )
# plot_ordination(loriSandyHealthy, ord,  color = "Username", shape = "Site")# label = "Label")

#ord <- ordinate(subset_samples(c.people, Site == "gut"), method = "DCA", distance = "bray" )
make_ord_for(c.people, color = "Username")


```

### SJ

His doctor says he has high _Klebsiella_, but how high is high? Klebsiella pneumoniae

```{r checkKelbsiella}
# 
# mhg_density_for_taxa(people.species,"Klebsiella pneumoniae") + 
#   annotate(
#   geom = "segment",
#   color = "blue",
#   x = as.numeric(otu_table(
#   subset_samples(sprague.species.norm, Site=="gut"))["Klebsiella pneumoniae"] / 10000),
#   xend = as.numeric(otu_table(
#   subset_samples(sprague.species.norm,Site == "gut"))["Klebsiella pneumoniae"] / 10000),
#   y = 0,
#   yend = 0.5
#   )
# 

```



### Mr. C

```{r}

conor.phylum <- subset_samples(people.phylum,  Username == c("em3"))
conor  <- subset_samples(people.norm,  Username == c("em3"))

c.people <- subset_samples(people.family, Site == "gut") # compare set
sample_data(c.people)$Username <- "Other"

sample_data(c.people)$Username[sample_data(c.people)$SSR == 12356] <- "Conor"


make_ord_for(c.people)

```

### FF

Here's a person from Norway who has long suffered from IBS.

```{r}

ff.phylum <- subset_samples(people.phylum,  Username == c("ff"))
ff  <- subset_samples(people.norm,  Username == c("ff"))

ff.people <- subset_samples(people.family, Site == "gut" & Condition == "Healthy") # compare set
ff.people <- merge_phyloseq(subset_samples(people.family,Username=="ff"),ff.people)
sample_data(ff.people)$Username <- "Other"

sample_data(ff.people)$Username[sample_data(ff.people)$SSR == 379847] <- "FF"


make_ord_for(ff.people)

ff.people.genus <- subset_samples(people.norm, Site == "gut" & Condition == "Healthy") # compare set
ff.people.genus <- merge_phyloseq(subset_samples(people.norm,Username=="ff"),ff.people.genus)

mhg_density_for_taxa(ff.people.genus,"Sutterella") + 
  annotate(
  geom = "segment",
  color = "blue",
  x = as.numeric(otu_table(
  subset_samples(ff.people.genus, Username == "ff"))["Sutterella"] / 10000),
  xend = as.numeric(otu_table(
  subset_samples(ff.people.genus, Username == "ff"))["Sutterella"] / 10000),
  y = -Inf,
  yend = Inf
  )


```


```{r}
mhg_density_for_taxa(people.norm,"Fusicatenibacter") + 
  annotate(
  geom = "segment",
  color = "blue",
  x = as.numeric(otu_table(
  subset_samples(people.norm, Username == "em3"))["Fusicatenibacter"] / 10000),
  xend = as.numeric(otu_table(
  subset_samples(people.norm, Username == "em3"))["Fusicatenibacter"] / 10000),
  y = -Inf,
  yend = Inf
  )
```

## High Proteobacteria

Usually I associate the phylum _Proteobacteria_ with trouble. So when recently somebody showed me results with nearly 10%, I immediately wanted to know more.

```{r}
db1.phylum <- prune_taxa(taxa_sums(subset_samples(people.phylum, Username == "db1")) > 50, (subset_samples(people.phylum, Username == "db1")))
db1.genus <- prune_taxa(taxa_sums(subset_samples(people.norm, Username == "db1")) > 50, (subset_samples(people.norm, Username == "db1")))
db1.family <- prune_taxa(taxa_sums(subset_samples(people.family, Username == "db1")) > 50, (subset_samples(people.family, Username == "db1")))
# mhg_taxa(db1.genus,"Akkermansia")

kable(mhg_abundance(db1.genus)/10000,col.names = c("db1 %"))
```

Most of the _Proteobacteria_ is from the genus _Parasutterella_.  I'm not sure what this one does, but the levels in this sample are much higher than anything I've seen before.

```{r caseStudyGutIssuesPeopleOrdLiza}
people.tmp <- subset_samples(people.family, Site=="gut")
ssr.tmp <- sample_data(subset_samples(people.tmp, Username == "db1"))$SSR

sample_data(people.tmp)$Username <- "People"

sample_data(people.tmp)$Username[sample_data(people.tmp)$SSR == ssr.tmp] <- "Liza"
make_ord_for(people.tmp)

```


