## Gut Issues

NOTE: THIS CHAPTER IS AN EARLY DRAFT


Probably the most common reason people are motivated to study their microbiome is gut dysbiosis. The word "dysbiosis" is as good a diagnosis as any for a problem that is often frustratingly difficult to pin down.

Ten years ago, CT was in excellent health, active in sports with a lean, muscular build. But after a trip to China, and a bout of what he thought was food poisoning, he hasn't been the same. Food allergies, sensitivities to chemicals and bright lights, fatigue, and unexplained muscular problems. Doctors were baffled. Nothing seemed to help, and it looked like he might simply have to accept that this was the new way he'd have to live.

The first thing to note is that this sample is pretty similar in overall diversity to the healthy samples in our database. 

```{r CTOverview, fig.cap="The blue line is CT."}
ct.gut.family <- subset_samples(people.family, Username == "ct")

estimate_richness(ct.gut.family, measures = "Shannon")
estimate_richness(people.family.healthy.gut, measures = "Shannon") %>% pull(Shannon) %>% median()

estimate_richness(people.family.healthy.gut, measures = "Shannon") %>% arrange(Shannon) %>%   ggplot(aes(Shannon)) + geom_histogram(aes(x=Shannon), bins = 50) + 
  geom_vline(xintercept=estimate_richness(ct.gut.family, measures = "Shannon") %>% pull(),
             color = "blue")
```



CB is a health-conscious woman who has long suffered from gut issues.

Let's start with a heatmap of all her gut samples to see if we can spot any patterns

```{r caseStudyCBExplore}

cb.gut.genus <- subset_samples(people.gut.norm, Username == "cb")
cb.gut.genus <- prune_taxa(taxa_sums(cb.gut.genus)>42, cb.gut.genus)
plot_heatmap(cb.gut.genus, sample.order = "Date", sample.label = "Date")
# ggplot(data = sample_data(cb.gut.genus)[,c("Date","Reads")], aes(x=Date,y=Reads)) + geom_bar(stat="identity") 


```

Hard to tell from this single case. How about if we look at another person ("NJ1")


First let's make a similarity plot it to see if we can spot something unusual:

```{r}
#plot_top_heat(nj1)
people.tmp <- people.family
sample_data(people.tmp)$Username <-ifelse(sample_data(people.tmp)$Username=="NJ1","NJ1","People")
make_ord_for(subset_samples(people.tmp, Site == "gut"))
```

Nothing special there. She seems firmly in the middle of most of the rest of the people.

Now let's look for specific microbes at the phylum level:

```{r}

nj1 <- subset_samples(people.norm, Username == "NJ1")
nj1 <- prune_taxa(taxa_sums(nj1)>42, nj1)
nj1.phylum <- subset_samples(people.phylum, Username == "NJ1")
nj1.phylum <- prune_taxa(taxa_sums(nj1.phylum)>42, nj1.phylum)

knitr::kable(mhg_abundance(nj1.phylum)/10000, digits = 2)

```

Nothing obviously different there than with others I've seen. Might be nice to see more Verrucomicrobia in the second (7-25) sample, but nothing pops as particularly special. How about the genus level:

```{r, results = "asis"}
knitr::kable(mhg_abundance(nj1)/10000, digits = 2)

```

The most striking finding is how different the second sample is from the first. Why?

When I look closer, it appears that much of the difference is driven by one genus, _Colinsella_, of which she has among the highest abundances I've seen. In my own gut samples, I've never had more than 5% of that genus, but she has more than 8%. Is that good or bad? Who knows? But it's something I would spend more time studying.


```{r}

mhg_density_for_taxa(people.norm, "Collinsella") +
  annotate(
  geom = "segment",
  color = "blue",
  x = as.numeric(otu_table(
  subset_samples(people.norm, Username == "NJ1")
  )["Collinsella"]) / 10000,
  xend = as.numeric(otu_table(
  subset_samples(people.norm, Username == "NJ1")
  )["Collinsella"]) / 10000 ,
  y = -Inf,
  yend = Inf
  )

           
```




Clearly different at nearly all taxa.  Which microbe disappeared in the second sample?

```{r}

psmr::unique_taxa(subset_samples(nj1, Date > "2017-07-01"), subset_samples(nj1, Date < "2017-07-01"))
```

and vice versa:

```{r}
kable(psmr::unique_taxa(subset_samples(nj1, Date < "2017-07-01"),subset_samples(nj1, Date > "2017-07-01"))[1]/10000, digits = 2, col.names = c("% Unique to first sample"))
```


Let's look at another case:  "Ed" is a German man on a mostly vegetarian diet.

```{r}

ed <- subset_samples(people.norm, Username == "AD" & Site == "gut")
ed.norm <- prune_taxa(taxa_sums(ed)>42, ed)

#plot_top_heat(ed.norm)

c.people <- people.norm # compare set
sample_data(c.people)$Username <- "Other"


ed.all <- subset_samples(people.phylum, Username == "AD")
ed.ssr <- sample_data(ed.all)$SSR
sample_data(c.people)$Username[sample_data(c.people)$SSR %in% ed.ssr] <- "You"


  
  
ed.phylum <- subset_samples(people.phylum, Username == "AD" & Site == "gut")
ed.phylum.norm <- prune_taxa(taxa_sums(ed.phylum)>42, ed.phylum)
mhg_abundance(ed.phylum.norm)/10000

mhg_abundance(ed.norm)/10000

# using another database: not for public
# load(paste0(DATA_DIR,"/ubiome corp/healthy/healthy897Genus.RData"))
# sample_data(healthy.ps)$Site <- "gut"
# healthyPlusEd <- merge_phyloseq(healthy.ps,ed.norm)
# ord <- ordinate(healthyPlusEd, method = "NMDS", distance = "bray" )
# plot_ordination(healthyPlusEd, ord,  color = "Username", shape = "Site")# label = "Label")


#ord <- ordinate(subset_samples(c.people, Site == "gut"), method = "DCA", distance = "bray" )

make_ord_for(subset_samples(c.people, Site == "gut"), color = "Username")

```

### Celiac Patients


L and her friend S suffer from the gluten disorder called Celiac disease.

```{r}
l.phylum <- subset_samples(people.phylum,  Username == c("em1"))
l  <- subset_samples(people.norm,  Username == c("em1"))
s.phylum <- subset_samples(people.phylum,  Username == c("em2"))
s <-subset_samples(people.norm,  Username == c("em2"))
mhg_abundance(merge_phyloseq(l.phylum,s.phylum),top=6)/10000

mhg_abundance(merge_phyloseq(l,s),top=10)/10000

```


I don't often see L's level of Fusobacterium. A quick check, though, shows that she's not the highest.

```{r}
mhg_density_for_taxa(people.norm,"Fusobacterium") + 
  annotate(
  geom = "segment",
  color = "blue",
  x = as.numeric(otu_table(
  subset_samples(people.norm, Username == "em1"))["Fusobacterium"] / 10000),
  xend = as.numeric(otu_table(
  subset_samples(people.norm, Username == "em1"))["Fusobacterium"] / 10000),
  y = -Inf,
  yend = Inf
  )

plot_richness(merge_phyloseq(subset_samples(people.family,Username=="em1"),subset_samples(people.family,Username=="em2")), measures=c("InvSimpson"))

ls.ssr <- c(sample_data(l)$SSR,sample_data(s)$SSR)
  
  
c.people <- subset_samples(people.family, Site == "gut") # compare set
sample_data(c.people)$Username <- "Other"

#sample_data(c.people)$Username[sample_data(c.people)$SSR %in% ls.ssr] <- subset_samples(c.people)

sample_data(c.people)$Username[sample_data(c.people)$SSR %in% ls.ssr] <- sample_data(people.norm)$Username[sample_data(people.norm)$SSR %in% ls.ssr]

# loriSandyHealthy <- merge_phyloseq(lori,sandy,healthy.ps)
# ord <- ordinate(loriSandyHealthy, method = "NMDS", distance = "bray" )
# plot_ordination(loriSandyHealthy, ord,  color = "Username", shape = "Site")# label = "Label")

#ord <- ordinate(subset_samples(c.people, Site == "gut"), method = "DCA", distance = "bray" )
make_ord_for(c.people, color = "Username")


```

### SJ

His doctor says he has high _Klebsiella_, but how high is high? 

Here are the genus-level microbes that are out of range:


```{r checkSJgenus}


sj.genus <- subset_samples(people.norm,Username == "sj")



mhg_outlier(mhg_range_for(sj.genus),mhg_range_for(gut.norm)) %>% filter(original>max)


```


and species:

```{r checkSJSpecies}
sj.species <- subset_samples(people.species,Username == "sj")
mhg_outlier(mhg_range_for(sj.species),mhg_range_for(sprague.species)) %>% filter(original>max)
# 
# mhg_density_for_taxa(people.species,"Klebsiella pneumoniae") + 
#   annotate(
#   geom = "segment",
#   color = "blue",
#   x = as.numeric(otu_table(
#   subset_samples(sprague.species.norm, Site=="gut"))["Klebsiella pneumoniae"] / 10000),
#   xend = as.numeric(otu_table(
#   subset_samples(sprague.species.norm,Site == "gut"))["Klebsiella pneumoniae"] / 10000),
#   y = 0,
#   yend = 0.5
#   )
# 

```



### Mr. C

```{r}

c.phylum <- subset_samples(people.phylum,  Username == c("em3"))
c  <- subset_samples(people.norm,  Username == c("em3"))

c.people <- subset_samples(people.family, Site == "gut") # compare set
sample_data(c.people)$Username <- "Other"

sample_data(c.people)$Username[sample_data(c.people)$SSR == 12356] <- "Conor"


make_ord_for(c.people)

```

### FF

Here's a person from Norway who has long suffered from IBS.

```{r}

ff.phylum <- subset_samples(people.phylum,  Username == c("ff"))
ff  <- subset_samples(people.norm,  Username == c("ff"))

ff.people <- subset_samples(people.family, Site == "gut" & Condition == "Healthy") # compare set
ff.people <- merge_phyloseq(subset_samples(people.family,Username=="ff"),ff.people)
sample_data(ff.people)$Username <- "Other"

sample_data(ff.people)$Username[sample_data(ff.people)$SSR == 379847] <- "FF"


make_ord_for(ff.people)

ff.people.genus <- subset_samples(people.norm, Site == "gut" & Condition == "Healthy") # compare set
ff.people.genus <- merge_phyloseq(subset_samples(people.norm,Username=="ff"),ff.people.genus)

mhg_density_for_taxa(ff.people.genus,"Sutterella") + 
  annotate(
  geom = "segment",
  color = "blue",
  x = as.numeric(otu_table(
  subset_samples(ff.people.genus, Username == "ff"))["Sutterella"] / 10000),
  xend = as.numeric(otu_table(
  subset_samples(ff.people.genus, Username == "ff"))["Sutterella"] / 10000),
  y = -Inf,
  yend = Inf
  )


```


```{r}
mhg_density_for_taxa(people.norm,"Fusicatenibacter") + 
  annotate(
  geom = "segment",
  color = "blue",
  x = as.numeric(otu_table(
  subset_samples(people.norm, Username == "em3"))["Fusicatenibacter"] / 10000),
  xend = as.numeric(otu_table(
  subset_samples(people.norm, Username == "em3"))["Fusicatenibacter"] / 10000),
  y = -Inf,
  yend = Inf
  )
```

### High Proteobacteria

Usually I associate the phylum _Proteobacteria_ with trouble. So when recently somebody showed me results with nearly 10%, I immediately wanted to know more.

```{r}
db1.phylum <- prune_taxa(taxa_sums(subset_samples(people.phylum, Username == "db1")) > 50, (subset_samples(people.phylum, Username == "db1")))
db1.genus <- prune_taxa(taxa_sums(subset_samples(people.norm, Username == "db1")) > 50, (subset_samples(people.norm, Username == "db1")))
db1.family <- prune_taxa(taxa_sums(subset_samples(people.family, Username == "db1")) > 50, (subset_samples(people.family, Username == "db1")))
# mhg_taxa(db1.genus,"Akkermansia")

kable(mhg_abundance(db1.genus)/10000,col.names = c("db1 %"))
```

Most of the _Proteobacteria_ is from the genus _Parasutterella_.  I'm not sure what this one does, but the levels in this sample are much higher than anything I've seen before.

```{r caseStudyGutIssuesPeopleOrdHH}
people.tmp <- subset_samples(people.family, Site=="gut")
ssr.tmp <- sample_data(subset_samples(people.tmp, Username == "db1"))$SSR

sample_data(people.tmp)$Username <- "People"

sample_data(people.tmp)$Username[sample_data(people.tmp)$SSR == ssr.tmp] <- "HH"
make_ord_for(people.tmp)

```


